---
title: "reactR"
author: "Ludvig Larsson"
date: "2022-09-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(reactR)
```

Create R package

```{r}

usethis::create_package("~/react_r_apps/featureviewer/")

```

Inject the widget templating

Dependencies:

1. openseadragon ^3.1.0"
2. paper ^0.12.17
3. d3 ^7.6.1

```{r}

withr::with_dir(
  "~/react_r_apps/ftrviewer/",
  reactR::scaffoldReactWidget("ftrviewer", list("openseadragon" = "^3.1.0",
                                                "paper" =  "^0.12.17",
                                                "d3" = "^7.6.1"), edit = FALSE)
)

```

The next step is to navigate to the newly-created paper project and run the following R commands:

```{r}

getwd()

system("yarn install")
system("yarn run webpack")

```

Now that the widget’s JavaScript is compiled, go ahead and install the R package:

```{r}

devtools::document()
devtools::install(quick = TRUE)

```

Now that the widget’s JavaScript is compiled, and the R package is installed, run app.R to see a demo in action:

```{r}

shiny::runApp()

```

Go for a spin:

```{r warning=FALSE, message=FALSE}

system("yarn install")
system("yarn cache clean")
system("yarn run webpack")
devtools::document()
devtools::install()
library(ftrviewer)

```

```{r}
ftrviewer()
```

