---
title: "reactR"
author: "Ludvig Larsson"
date: "2022-09-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# JS code

The actual javascript code can be found in the `srcjs` folder. Here, you can find code for React components, CSS and other utilities.

# Create R code

This code is adapted from the `reactR` tutorial.

```{r}
library(reactR)
```

Create R package

```{r}

usethis::create_package("~/react_r_apps/featureviewer/")

```

Inject the widget templating

Dependencies:

1. openseadragon ^3.1.0"
2. paper ^0.12.17
3. d3 ^7.6.1

```{r}

withr::with_dir(
  "~/react_r_apps/ftrviewer/",
  reactR::scaffoldReactWidget("ftrviewer", list("openseadragon" = "^3.1.0",
                                                "paper" =  "^0.12.17",
                                                "d3" = "^7.6.1"), edit = FALSE)
)

```

The next step is to navigate to the newly-created paper project and run the following R commands:

```{r}

getwd()

system("yarn install")
system("yarn run webpack")

```

Now that the widget’s JavaScript is compiled, go ahead and install the R package:

```{r}

devtools::document()
devtools::install(quick = TRUE)

```

Now that the widget’s JavaScript is compiled, and the R package is installed, run app.R to see a demo in action:

```{r}

shiny::runApp()

```

Go for a spin:

```{r warning=FALSE, message=FALSE}

system("yarn install")
system("yarn cache clean")
system("yarn run webpack")
devtools::document()
devtools::install()
library(ftrviewer)

```

```{r}
ftrviewer()
```

# Comments

In the `app.js` file we can see that a React component called `App` is defined which uses the `OSG` component. The `OSG` component inherits props from `App` which in turn receives input from the R function `ftrviewer` defined in `R/ftrviewer.R`.

The functions in `R/ftrviewer.R` were created using reactR and do not need to be modified, except that you need to specify the correct package name in `ftrviewer()` and `ftrviewerOutput()`. `scaffoldReactWidget` automatically creates a full R package so it can be a good idea to give it a unique name, s.a. `ftrviewer` in this case. However, when we move the code to `semla`, we need to change make these changes. I have commented the two lines of code.

If you need to change the widget `ftrviewer`, you'll have to modify the JS code. If we want to pass more parameters from the `ftrviewer()` R function to the React app, you can add additional arguments. Let's say we add a new argument called `spotsize`, we should the be able to access it from `props.spotsize` in the `App` React component.

## Window size

I have set the container size to be static which means that it will not be rescaled when the browser window is changed. This size is controlled with `container_width/container_height`. This was necessary to avoid a bunch of scaling issues in the app and I would highly recommend keeping it this way.

## File server (host + port)

Use `host+port` to define the address to the static file server created with the `semla` function `file_server()`.

## SampleID

Integer specifying the sample ID. Used to select image tiles and data for selected sample.

## scaleByOpacity

Logical which can be used to scale spot opacity. Same as `scale_alpha` in `MapFeatures`.

## isNumeric

Logical used to switch between numeric and categorical data.

## Levels

Levels for categorical variables used to map colors