---
title: "Create dataset"
date: "Last compiled: `r format(Sys.time(), '%d %B %Y')`"
format: html
editor: visual
editor_options: 
  chunk_output_type: inline
---

In this notebook, we'll have a look at how you can create a `Seurat` object compatible with `semla`.

## Load libraries

```{r message=FALSE, warning=FALSE}
library(semla)
library(magick)
library(jsonlite)
library(tibble)
```

## Load coordinates

The `Staffli` object requires image and coordinate data that can be retireved from the spaceranger output files.

Below we'll use the mouse brain and mouse colon example data:

```{r}
he_imgs <- c(system.file("extdata/mousebrain", 
                         "spatial/tissue_hires_image.jpg", 
                          package = "semla"),
             system.file("extdata/mousecolon", 
                         "spatial/tissue_hires_image.jpg", 
                         package = "semla"))
spotfiles <- c(system.file("extdata/mousebrain", 
                           "spatial/tissue_positions_list.csv", 
                           package = "semla"),
               system.file("extdata/mousecolon", 
                           "spatial/tissue_positions_list.csv", 
                           package = "semla"))
jsonfiles <- c(system.file("extdata/mousebrain", 
                           "spatial/scalefactors_json.json", 
                           package = "semla"),
               system.file("extdata/mousecolon", 
                           "spatial/scalefactors_json.json", 
                           package = "semla"))
```

When loading coordinates from multiple samples, you need to change the barcode IDs so that the suffix matches their respective sampleID. For instance, a barcode in sample 1 might be called CATACAAAGCCGAACC-1 and in sample 2 the same barcode should be CATACAAAGCCGAACC-2. The `coordinates` tibble below contains the barcode IDs, the pixel coordinates and a column with the sampleIDs.

```{r eval=FALSE}
# Read coordinates
coordinates <- do.call(bind_rows, lapply(seq_along(spotfiles), function(i) {
  sample_coords <- read.csv(spotfiles[i], header = FALSE) |> 
    as_tibble() |>
    setNames(nm = c("barcode", "selected", "y", "x", "pxl_row_in_fullres", "pxl_col_in_fullres")) |> 
    mutate(sampleID = i) |> # Add a unique sampleID
    filter(selected == 1) |> 
    select(barcode, pxl_col_in_fullres, pxl_row_in_fullres, sampleID) |> 
    mutate(barcode = gsub(pattern = "-\\d*", # Replace barcode suffix with sampleID
                          replacement = paste0("-", i), 
                          x = barcode))
  return(sample_coords)
}))
```

`semla` provides `LoadSpatialCoordinates` to make the task a little easier:

```{r}
coordinates <- LoadSpatialCoordinates(coordinatefiles = spotfiles, verbose = FALSE) |> 
  select(barcode, pxl_row_in_fullres, pxl_col_in_fullres, sampleID)

coordinates

# Check number of spots per sample
table(coordinates$sampleID)
```


## Fetch image info

Next, we'll fetch meta data for the sample H&E images. Here, the tibble must contain the width and height of `he_imgs` as well as a sampleID column.

```{r}
# Create image_info
image_info <- do.call(bind_rows, lapply(seq_along(he_imgs), function(i) {
  he_imgs[i] |> 
    image_read() |> 
    image_info() |> 
    mutate(sampleID = paste0(i))
}))
image_info
```

## Define image dimensions

The `jsonfiles` contain scaling factors that allows us to find the original H&E image dimensions. We can load the scale factors with `read_json` and store them in a tibble. The tibble should also contain a sampleID column.

```{r}
# Read scalefactors
scalefactors <- do.call(bind_rows, lapply(seq_along(jsonfiles), function(i) {
  jsonlite::read_json(jsonfiles[i]) |> 
    as_tibble() |> 
    mutate(sampleID = paste0(i))
}))
scalefactors
```

Once we have the scale factors, we can add additional columns to `image_info` which specify the dimensions of the original H&E images:

```{r}
# Add additional columns to image_info using scale factors
image_info <- image_info |> 
  mutate(full_width = round(width/scalefactors$tissue_hires_scalef[row_number()]),
         full_height = round(height/scalefactors$tissue_hires_scalef[row_number()])) |> 
  mutate(type = "tissue_hires") |> 
  select(format, width, height, full_width, full_height, 
         colorspace, filesize, density, sampleID, type)
image_info
```
Finally, we are ready to create the `Staffli` object. Here we'll provide the `he_imgs` so that we can load the images later with `LoadImages`. The `coordinates` are stored in the `meta_data` slot. `image_info` will be used by the plot functions provided in `semla` to define the dimensions of the plot area.

```{r}
# Create Staffli object
staffli_object <- CreateStaffliObject(imgs = he_imgs, 
                                      meta_data = coordinates, 
                                      image_info = image_info, 
                                      scalefactors = scalefactors)
staffli_object
```

## Use a `Staffli` object in `Seurat`

Now let's see how we can incorporate our `Staffli` object in `Seurat`. First, we'll load the gene expression matrices for our samples and merge them. 

```{r}
# Get paths for expression matrices
expr_matrix_files <- he_imgs <- c(system.file("extdata/mousebrain", 
                         "filtered_feature_bc_matrix.h5", 
                          package = "semla"),
             system.file("extdata/mousecolon", 
                         "filtered_feature_bc_matrix.h5", 
                         package = "semla"))
```

Before merging the matrices, it's important to rename the barcodes so that they match the barcodes in `Staffli_object`.

```{r eval=FALSE}
exprMatList <- lapply(seq_along(expr_matrix_files), function(i) {
  exprMat <- Seurat::Read10X_h5(expr_matrix_files[i])
  colnames(exprMat) <- gsub(pattern = "-\\d*", # Replace barcode suffix with sampleID
                          replacement = paste0("-", i), 
                          x = colnames(exprMat))
  return(exprMat)
})

# Merge expression matrices
mergedExprMat <- SeuratObject::RowMergeSparseMatrices(exprMatList[[1]], exprMatList[[2]])

```

Alternatively, use the `LoadAndMergeMatrices` function from `semla`:

```{r warning=FALSE, message=FALSE}
mergedExprMat <- LoadAndMergeMatrices(samplefiles = expr_matrix_files, verbose = FALSE)

# Create Seurat object
se <- CreateSeuratObject(counts = mergedExprMat)
```

Rearrange `Staffli` object to match `Seurat` object. If the barcode are mismatched, make sure to subset the two object to contain intersecting barcodes.

```{r}
staffli_object@meta_data <- staffli_object@meta_data[match(colnames(se), staffli_object@meta_data$barcode), ]
```

Check that the `Staffli` object matches `Seurat` object:

```{r}
all(staffli_object@meta_data$barcode == colnames(se))
```

Place `Staffli` object in the `Seurat` object:

```{r}
se@tools$Staffli <- staffli_object
```

Now the `semla` functions should be compatible with the `Seurat` object!

```{r}
MapFeatures(se, features = "nFeature_RNA")

se <- LoadImages(se, verbose = FALSE)
MapFeatures(se, features = "nFeature_RNA", image_use = "raw", override_plot_dims = TRUE)
```


<br>

------------------------------------------------------------------------

<details open>

<summary>**Package version**</summary>

-   `semla`: `r packageVersion("semla")`

</details>

<details>

<summary>**Session info**</summary>

```{r}
sessionInfo()
```

</details>

<br>
