---
title: "Cell type mapping with NNLS - comparison"
date: "Last compiled: `r format(Sys.time(), '%d %B %Y')`"
format: html
editor: visual
editor_options: 
  chunk_output_type: inline
---

```{r, message=FALSE, warning=FALSE}
library(semla)
library(tibble)
library(dplyr)
if (!requireNamespace('TabulaMurisSenisData', quietly = TRUE)) {
  BiocManager::install("TabulaMurisSenisData")
}

library(TabulaMurisSenisData)
library(SingleCellExperiment)
library(tidyr)
library(patchwork)
library(ggplot2)
```

```{r include=FALSE}

se_allen <- readRDS("~/10x_data/R_objects/allen_brain")
se_brain_spatial <- readRDS("~/10x_data/R_objects/se_brain_spatial")
se_kidney_singlecell <- readRDS("~/10x_data/R_objects/se_kidney_singlecell")
se_kidney_spatial <- readRDS("~/10x_data/R_objects/se_kidney_spatial")

```

```{r include=FALSE, eval=FALSE}

var_genes <- FindVariableFeatures(se_kidney_spatial, nfeatures = 5000) |> VariableFeatures()
writeLines(text = var_genes, con = "../stereoscope/mKidney/variable_genes.txt")

umis_spatial <- GetAssayData(se_kidney_spatial, slot = "counts")
umis_spatial <- umis_spatial[rowSums(umis_spatial) > 0, ]

data.table::fwrite(umis_spatial |> t() |> data.frame(), file = "../stereoscope/mKidney/ST_section_1.tsv", quote = FALSE, sep = "\t", row.names = TRUE, col.names = TRUE)

set.seed(123)
mData <- se_kidney_singlecell@meta.data |> 
  rownames_to_column(var = "barcode") |> 
  group_by(free_annotation) |> 
  filter(row_number() %in% sample(x = 1:n(), size = ifelse(n() > 250, 250, n()))) |> 
  filter(n() >= 10)

umis_sc <- GetAssayData(se_kidney_singlecell, slot = "counts")[, mData$barcode]
umis_sc <- umis_sc[rowSums(umis_sc) > 0, ]

data.table::fwrite(umis_sc |> t() |> data.frame(), file = "../stereoscope/mKidney/exprMat.tsv", quote = FALSE, sep = "\t", row.names = TRUE, col.names = TRUE)

write.table(mData |> dplyr::select(barcode, free_annotation) |> 
              dplyr::rename(celltype = free_annotation) |> 
              mutate(celltype = gsub(pattern = "Epcam +|CD45 +|Pecam +|Stroma +", 
                                     replacement = "", x = celltype)) |> 
              mutate(celltype = gsub(pattern = " ", replacement = "_", x = celltype)) |> 
              data.frame(row.names = 1), 
            file = "../stereoscope/mKidney/mData.tsv", sep = "\t", row.names = TRUE, quote = FALSE)

```

### Comparison with stereoscope

If we compare the results from the NNLS cell type mapping method with the results from stereoscope, we can see that there is a high concordance between the results.

```{r include=FALSE}
se_brain_spatial[["celltypeprops"]] <- readRDS(file = "../vignettes/mousebrain/comparison/celltypeprops")
se_brain_spatial[["stereoscope"]] <- readRDS(file = "../vignettes/mousebrain/comparison/stereoscope")
```

```{r fig.asp=0.9, fig.width=4}
corMat <- cor(t(GetAssayData(se_brain_spatial, assay = "celltypeprops")) |> as.matrix(), 
              t(GetAssayData(se_brain_spatial, assay = "stereoscope")) |> as.matrix())
corMat <- corMat[intersect(rownames(corMat), colnames(corMat)), intersect(rownames(corMat), colnames(corMat))]

gg <- corMat |> 
  data.frame(check.names = FALSE) |> 
  rownames_to_column(var = "CellType_NNLS") |> 
  as_tibble() |> 
  tidyr::pivot_longer(all_of(rownames(corMat)), names_to = "CellType_stereoscope", 
                      values_to = "correlation")

ggplot(gg, aes(CellType_NNLS, CellType_stereoscope, fill = correlation)) +
  geom_tile() +
  scale_fill_gradientn(colours = RColorBrewer::brewer.pal(n = 11, name = "RdBu") |> rev(), limits = c(-1, 1)) +
  scale_y_discrete(limits = rev) +
  scale_x_discrete(position = "top") +
  theme(axis.text.x = element_text(angle = 60, hjust = 0))
```

Let's map the proportions on the H&E image and compare them side by side:

```{r}
plots <- lapply(seq_along(selected_celltypes), function(i) {
  DefaultAssay(se_brain_spatial) <- "celltypeprops"
  p1 <- MapFeatures(se_brain_spatial, pt_size = 1.3, ncol = 2,
            features = selected_celltypes[i], image_use = "raw",
            arrange_features = "row", scale = "shared", 
            override_plot_dims = TRUE,
            colors = RColorBrewer::brewer.pal(n = 9, name = "Spectral") |> rev(), 
            scale_alpha = TRUE) +
    plot_layout(guides = "collect") &
    labs(fill = "NNLS\nprop")
  DefaultAssay(se_brain_spatial) <- "stereoscope"
  p2 <- MapFeatures(se_brain_spatial, pt_size = 1.3, ncol = 2,
            features = selected_celltypes[i], image_use = "raw",
            arrange_features = "row", scale = "shared", 
            override_plot_dims = TRUE,
            colors = RColorBrewer::brewer.pal(n = 9, name = "Spectral") |> rev(), 
            scale_alpha = TRUE) +
    plot_layout(guides = "collect") &
    labs(fill = "stereoscope\nprop")
  p <- p1 / p2  &
  theme(legend.position = "right", legend.margin = margin(b = 50),
        legend.text = element_text(angle = 0), plot.title = element_text())
  return(p)
}) |> setNames(nm = selected_celltypes)
```

### NNLS vs stereoscope {.tabset}

```{r results='asis', echo=FALSE, fig.asp=1, fig.width=4}
for(i in 1:length(selected_celltypes)){
  cat("#### ", selected_celltypes[i],"\n")
  print(plots[[i]])
  cat("\n\n")
}
```

# Load kidney results

```{r}
propMat <- read.table("../stereoscope/mKidney/ST_section_1/W.2023-03-07105412.956337.tsv") |> t()


eMat <- matrix(ncol = ncol(se_kidney_spatial), nrow = nrow(propMat), 
               dimnames = list(rownames(propMat), colnames(se_kidney_spatial)))
shared_spots <- intersect(colnames(propMat), colnames(se_kidney_spatial))
eMat[, shared_spots] <- propMat[, shared_spots]
rownames(eMat) <- gsub(pattern = "_", replacement = " ", x = rownames(eMat))
se_kidney_spatial[["stereoscope"]] <- CreateAssayObject(data = eMat)

```

```{r}

se_kidney_singlecell$celltype <- se_kidney_singlecell@meta.data |> 
  mutate(celltype = gsub(pattern = "Epcam +|CD45 +|Pecam +|Stroma +", 
                                     replacement = "", x = free_annotation)) |> 
  pull(celltype)

DefaultAssay(se_kidney_spatial) <- "Spatial"
se_kidney_spatial <- RunNNLS(object = se_kidney_spatial, 
                      singlecell_object = se_kidney_singlecell, 
                      groups = "celltype")

# Check available cell types
rownames(se_kidney_spatial)
```


```{r include=FALSE}
se_kidney_spatial[["celltypeprops"]] <- 
  readRDS(file = "../vignettes/comparison/mousekidney/celltypeprops")
se_kidney_spatial[["stereoscope"]] <- 
  readRDS(file = "../vignettes/comparison/mousekidney/stereoscope")
```

```{r fig.asp=0.9, fig.width=4}
corMat <- cor(t(GetAssayData(se_kidney_spatial, assay = "celltypeprops")) |> as.matrix(), 
              t(GetAssayData(se_kidney_spatial, assay = "stereoscope")) |> as.matrix())
corMat <- corMat[intersect(rownames(corMat), colnames(corMat)), intersect(rownames(corMat), colnames(corMat))]

gg <- corMat |> 
  data.frame(check.names = FALSE) |> 
  rownames_to_column(var = "CellType_NNLS") |> 
  as_tibble() |> 
  tidyr::pivot_longer(all_of(rownames(corMat)), names_to = "CellType_stereoscope", 
                      values_to = "correlation")

ggplot(gg, aes(CellType_NNLS, CellType_stereoscope, fill = correlation)) +
  geom_tile() +
  scale_fill_gradientn(colours = RColorBrewer::brewer.pal(n = 11, name = "RdBu") |> rev(), limits = c(-1, 1)) +
  scale_y_discrete(limits = rev) +
  scale_x_discrete(position = "top") +
  theme(axis.text.x = element_text(angle = 60, hjust = 0))
```

Let's map the proportions on the H&E image and compare them side by side:

```{r}

se_kidney_spatial <- LoadImages(se_kidney_spatial)
plots <- lapply(1:nrow(se_kidney_spatial), function(i) {
  celltype <- rownames(se_kidney_spatial)[i]
  DefaultAssay(se_kidney_spatial) <- "celltypeprops"
  p1 <- MapFeatures(se_kidney_spatial, pt_size = 1.3, ncol = 1,
            features = celltype, image_use = "raw",
            arrange_features = "row", scale = "shared", 
            override_plot_dims = TRUE,
            colors = RColorBrewer::brewer.pal(n = 9, name = "Spectral") |> rev(), 
            scale_alpha = TRUE) +
    plot_layout(guides = "collect") &
    labs(fill = "NNLS\nprop")
  DefaultAssay(se_kidney_spatial) <- "stereoscope"
  p2 <- MapFeatures(se_kidney_spatial, pt_size = 1.3, ncol = 1,
            features = celltype, image_use = "raw",
            arrange_features = "row", scale = "shared", 
            override_plot_dims = TRUE,
            colors = RColorBrewer::brewer.pal(n = 9, name = "Spectral") |> rev(), 
            scale_alpha = TRUE) +
    plot_layout(guides = "collect") &
    labs(fill = "stereoscope\nprop")
  p <- p1 - p2  &
  theme(legend.position = "right", legend.margin = margin(b = 50),
        legend.text = element_text(angle = 0), plot.title = element_text())
  return(p)
}) |> setNames(nm = rownames(se_kidney_spatial))
```

### NNLS vs stereoscope mouse kidney {.tabset}

```{r results='asis', echo=FALSE, fig.width=4}
for(i in 1:nrow(se_kidney_spatial)){
  cat("#### ", rownames(se_kidney_spatial)[i],"\n")
  print(plots[[i]])
  cat("\n\n")
}
```