---
title: "Get started"
format: 
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The goal with this package is to provide simple, fast and flexible methods for processing, analysis and visualization of spatially resolved transcriptomics data. Currently, the package only supports [Seurat](https://satijalab.org/seurat/) objects.

## Installation

`semla` is currently only available as a development version on GitHub:

```{r eval=FALSE}
remotes::install_github("ludvigla/semla")
```

Load library

```{r eval=FALSE}
library(semla)
packageVersion("semla")
```

## Authors

The package is written by *Ludvig Larsson* and *Lovisa Franzén* who are both part of the [Spatial Research lab](https://www.spatialresearch.org/) at [SciLifeLab](https://www.scilifelab.se/) in Stockholm. It is moreover maintained by *Javier Escudero Morlanes*.

## Publication

If you use `semla` in published research, please cite the following article:

> Larsson L, Franzén L, Ståhl PL, Lundeberg J. Semla: a versatile toolkit for spatially resolved transcriptomics analysis and visualization. *Bioinformatics*. 2023;39(10):btad626. doi:10.1093/bioinformatics/btad626

<br>
