---
title: "Get started"
format: 
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The goal with this package is to provide sample, fast and flexible methods for processing, analysis and visualization of spatially resolved transcriptomics data. Currently, the package only supports [Seurat](https://satijalab.org/seurat/) objects.

## Installation

`STUtility2` is currently only available as a development version on GitHub:

```{r eval=FALSE}
remotes::install_github("ludvigla/STUtility2")
```

Load library

```{r eval=FALSE}
library(STUtility2)
packageVersion("STUtility2")
```

## Authors

The package is written by *Ludvig Larsson* and *Lovisa Franzén* who are both part of the [Spatial Research lab](https://www.spatialresearch.org/) at [SciLifeLab](https://www.scilifelab.se/) in Stockholm.

## Publication

If you use `STUtility2` in published research, please cite the following article:
