---
title: "Subset and merge"
format: 
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    theme:
      light: flatly
      dark: darkly
editor: visual
---

Subsetting and merging data is done using the two functions `SubsetSTData` and `MergeSTData`.

If you use the generic functions `subset` and `merge`, these will work fine on `Seurat` objects but they will not be able to handle the spatial data that has been placed inside the `Seurat` object.

Let's load an example mouse brain 10x Visium data:

```{r warning=FALSE, message=FALSE}
library(STUtility2)
se_mcolon <- readRDS(system.file("extdata/mousecolon", 
                                 "se_mcolon", 
                                 package = "STUtility2"))
```

### Subset by selecting spots

```{r}
spots_to_keep <- c("AAACAAGTATCTCCCA-1", "AAACACCAATAACTGC-1", 
                   "AAACATTTCCCGGATT-1", "AAACCCGAACGAAATC-1", 
                   "AAACCGGGTAGGTACC-1", "AAACCGTTCGTCCAGG-1")

# Subset using selected spots
se_mcolon_small <- SubsetSTData(se_mcolon, spots = spots_to_keep)
se_mcolon_small
```

### Subset by selecting features

```{r}
genes_to_keep <- c("Hbb-bs", "Hba-a1", "Hba-a2", "Hbb-bt", "Slc6a3", "Th")

# Subset using selected genes
se_mcolon_fewgenes <- SubsetSTData(se_mcolon, features = genes_to_keep)
se_mcolon_fewgenes
```

### Subset with an expression

If we want to subset data using information from the meta.data slot, it might be easier to use an expression instead.

```{r}
# Filte by number of unique genes
se_mcolon_filtered <- SubsetSTData(se_mcolon, expression = nFeature_Spatial > 20)
se_mcolon_filtered
```
