---
title: "Subset and merge"
date: "Last compiled: `r format(Sys.time(), '%d %B %Y')`"
format: html
editor: visual
editor_options: 
  chunk_output_type: inline
---

Subseting and merging data is done using the two functions `SubsetSTData()` and `MergeSTData()`.

If you use the generic functions `subset` and `merge`, these will work fine on `Seurat` objects but they will not be able to handle the spatial data that has been placed inside the `Seurat` object by `semla`, causing the spatial data to be lost or corrupted.

Let's load an example mouse brain 10x Visium data:

```{r warning=FALSE, message=FALSE}
library(semla)
se_mcolon <- readRDS(system.file("extdata/mousecolon", 
                                 "se_mcolon", 
                                 package = "semla"))
se_mcolon
```

```{r fig.asp=1, fig.width=4}
MapFeaturesSummary(se_mcolon, features = "nFeature_Spatial", subplot_type = "histogram")
```

### Subset by selecting spots

The data can be subseted by specifying which spots to keep using their barcode ID.

```{r}
spots_to_keep <- c("AAACAAGTATCTCCCA-1", "AAACACCAATAACTGC-1", 
                   "AAACATTTCCCGGATT-1", "AAACCCGAACGAAATC-1", 
                   "AAACCGGGTAGGTACC-1", "AAACCGTTCGTCCAGG-1")

# Subset using selected spots
se_mcolon_small <- SubsetSTData(se_mcolon, spots = spots_to_keep)
se_mcolon_small
```

```{r fig.asp=1, fig.width=4}
MapFeaturesSummary(se_mcolon_small, features = "nFeature_Spatial", subplot_type = "histogram")
```

### Subset by selecting features

We can also extract data corresponding to a few selected genes of interest.

```{r}
genes_to_keep <- c("Hbb-bs", "Hba-a1", "Hba-a2", "Hbb-bt", "Slc6a3", "Th")

# Subset using selected genes
se_mcolon_fewgenes <- SubsetSTData(se_mcolon, features = genes_to_keep)
se_mcolon_fewgenes
```

```{r fig.asp=1, fig.width=4}
MapFeaturesSummary(se_mcolon_fewgenes, features = "nFeature_Spatial", subplot_type = "histogram")
```

### Subset with an expression

If we want to subset data using information from the `meta.data` slot, it might be easier to use an expression instead. This comes in handy when we want to perform QC filtering on our data.

```{r}
# Filter by number of unique genes
se_mcolon_filtered <- SubsetSTData(se_mcolon, expression = nFeature_Spatial > 30)
se_mcolon_filtered
```

```{r fig.asp=1, fig.width=4}
MapFeaturesSummary(se_mcolon_filtered, features = "nFeature_Spatial", subplot_type = "histogram")
```

### Merge two data sets

Finally, we can use `MergeSTData()` to join two objects. In this example, our colon data set will be merged with a brain data set.

```{r warning=FALSE, message=FALSE}
se_mbrain <- readRDS(system.file("extdata/mousebrain", 
                                 "se_mbrain", 
                                 package = "semla"))
```

```{r}
se_merged <- MergeSTData(se_mcolon, se_mbrain)

unique(se_merged$sample_id)
se_merged
```

```{r fig.width=6}
MapFeatures(se_merged, features = "nFeature_Spatial")
```

<br>

------------------------------------------------------------------------

<details open>

<summary>**Package version**</summary>

-   `semla`: `r packageVersion("semla")`

</details>

<details>

<summary>**Session info**</summary>

```{r}
sessionInfo()
```

</details>

<br>
